Surrogate model of aligned-spin binary neutron star waveforms using Gaussian process regression

### 2017-12-04 paper/paper.tex
Rewrote introduction. Added some more references.

%For the parameter most sensitive to waveform errors, the tidal parameter, we find that the error from the resulting surrogate leads to a systematic error in the measured tidal parameter $\tilde\Lambda$ of less than $\sim 50$ which is less than 10\% of typical tidal parameters.

@@

\pacs{
  % 04.25.D-,   % numerical relativity
  % 
  04.30.Db,   % gravitational wave generation and sources
  % 04.40.Dg,   % Relativistic stars: structure, stability, and oscillations
  % 04.70.Bw,   % classical black holes
  95.30.Sf,     % relativity and gravitation
  % 
  % 95.30.Lz,   % Hydrodynamics
  % 97.60.Jd    % Neutron stars
  % 97.60.Lf    % black holes (astrophysics)
  % 98.62.Mw    % Infall, accretion, and accretion disks
}

@@

\red{[[What is possible with BNS systems:]]}

The detection of the binary neutron star (BNS) system GW170817~\cite{GW170817} by the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors, has demonstrated that the sky position, distance, masses, spins, and tidal parameters can be measured with enough precision to answer fundamental physics questions such as identifying BNS mergers as a source of gamma ray bursts [[cite]], measuring the Hubble constant~\cite{GW170817Hubble}, and understanding the structure of neutron stars (NSs) through tidal interactions~\cite{GW170817}. Measuring these parameters accurately will also be necessary to properly interpret the gamma ray burst [[cite]] and optical afterglow [[cite]] that was found in conjunction. Furthermore, an estimate of the BNS merger rate of 320--4740 events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval)~\cite{GW170817} indicates that, when second generation detectors reach design sensitivity, including KAGRA~\cite{Somiya2012} and possibly LIGO-India~\cite{IyerSouradeepUnnikrishnan2011}, we could eventually observe a population of tens or hundreds of events after a couple of years of observation~\cite{LIGORate2010}. This will allow us to estimate the NS mass distribution [[cite]], spin distribution [[cite]], and universal equation of state (EOS) that is common to all NSs\cite{DelPozzoLiAgathos2013, LackeyWade2015}.

\red{[[Why we need more accurate waveform models to do these things:]]}

These results are obtained using a Bayesian analysis that compares the data to parameterized waveform templates. To avoid systematic errors highly accurate waveform templates are needed. In addition, the Bayesian parameter estimation alogrithms (Markov chain Monte Carlo and Nested Sampling [[parallel algorithms are becoming available]]) typically perform several million sequential waveform evaluations, requiring waveform evaluation times to be less than $\mathcal{O}(1s)$ in order to run in less than a month for each event.

%The primary quantities of interest for BNS merger observations are the sky position, masses, spins, and tidal parameters. These parameters, measured with Bayesian parameter estimation techniques, require highly accurate waveform models to avoid systematic errors. For example, post-Newtonian (PN) waveform models, which are truncated at the currently known 3.5PN order, lead to uncertainties in the waveform phase of approximately 10~rad up to a merger frequency of $\sim 1000$~Hz [[cite]]. This is approximately the same size as the contribution to the phase from tidal effects, and as a result, the systematic errors in the measured tidal parameter are as large as the tidal parameter itself [[cite]]. This also biases the recovered NS equation of state and radius [[cite]], as well as the masses and spins [[cite]].

\red{[[The problems with the current waveform models/templates.]]}

To date, there are a limited number BNS waveform models that include spin and tides, and that are fast enough for standard parameter estimation.  
TaylorF2: (not so good point particle. okay tidal terms)
For example, post-Newtonian (PN) waveform models, which are truncated at the currently known 3.5PN order, lead to uncertainties in the waveform phase of approximately 10~rad up to a merger frequency of $\sim 1000$~Hz [[cite]]. This is approximately the same size as the contribution to the phase from tidal effects, and as a result, the systematic errors in the measured tidal parameter are as large as the tidal parameter itself [[cite]]. This also biases the recovered NS equation of state and radius [[cite]], as well as the masses and spins [[cite]].

\*\_NRTides: (better/good point particle. better tidal terms)
Initial parameter estimation results for the first BNS event, GW170817 using a variety of waveform templates (TaylorF2, SEOBNRv4\_ROM\_NRTides, and IMRPhenomD\_NRTides) showed that there are noticeable differences in the measured parameters between the different waveform templates [[cite]], and in particular the tidal parameter. [[NR\_Tides is a fit to NR waveforms at high frequency and has only been fit to a limited number of low spin, low tidal parameter NR simulations.]]

TEOBResumS:

\red{[[What is good about the new EOB model:]]}

The effective one body (EOB) formalism, on the other hand, has significantly smaller uncertainties than PN waveforms. Variations within the EOB formalism lead to uncertainties in the waveform phase of $\sim 1$~rad [[cite or say as shown below]] as compared to PN waveform uncertainties of $\sim 10$~rad~\cite{HindererLackeyLangRead2010}. Furthermore, NR simulations of BNS systems now claim phase errors of less than 1~rad during the last $\sim 10$ orbits through merger [[cite]] and show good agreement with EOB models [[cite]].

Another model

\red{[[Problem with EOB (speed). Attempts to speed it up with optimization. Alternative using surrogates.]]}

Unfortunately, available implementations for generating EOB waveforms for BNS systems take anywhere from 10 minutes to a day to evaluate for a single waveform with a starting GW frequency of 10Hz [[cite]]. Bayesian parameter estimation algorithms such as Markov-chain Monte Carlo and nested sampling, however, typically require $10^7$--$10^8$ sequential waveform evaluations, making current EOB implementations unusable. Significant work has been done to optimize existing implementations of EOB for BBH systems [[EtienneMcWilliams]]. However, it is not clear that these optimizations can be readily adapted for the BNS model discussed here with dynamical tides. We therefore develop a surrogate model that accelerates the evaluation of EOB waveforms by interpolating between a sparsely sampled training set of waveforms within the parameter space.

Surrogate modeling techniques have had significant success in gravitational-wave data analysis for accelerating the evaluation of a waveform $h(t;\bx)$ given its waveform parameters $\bx$. The standard technique is to construct a training set of waveforms that sample the parameter space, then compress this training set by building a reduced, orthonormal basis whose linear combination can reproduce any waveform in the parameter space to some desired accuracy. This serves to interpolate the waveform as a function of time (or frequency for frequency-domain waveforms) in terms of a reduced set of coefficients corresponding to the reduced set of basis functions. In previous work, this reduced basis has been constructed either with singular value decomposition (SVD)~\cite{Puerrer2014, Puerrer2015} or with a greedy method that chooses a subset of the training-set waveforms that are as orthogonal to each other as possible, then orthonormalizes this subset with Graham-Schmidt orthogonalization~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}.

\red{[[Overview of surrogates. Typical methods that have been used. Their problems. What you are doing that is new.]]}

%The effective one body (EOB) formalism, on the other hand, has significantly smaller uncertainties than PN waveforms. Variations within the EOB formalism lead to uncertainties in the waveform phase of $\sim 1$~rad [[cite or say as shown below]] as compared to PN waveform uncertainties of $\sim 10$~rad. Furthermore, NR simulations of BNS systems now claim phase errors of less than 1~rad during the last $\sim 10$ orbits through merger [[cite]] and show good agreement with EOB models [[cite]]. Unfortunately, available implementations for generating EOB waveforms for BNS systems take anywhere from 10 minutes to a day to evaluate for a single waveform with a starting GW frequency of 10Hz [[cite]]. Bayesian parameter estimation algorithms such as Markov-chain Monte Carlo and nested sampling, however, typically require $10^7$--$10^8$ sequential waveform evaluations, making current EOB implementations unusable. Significant work has been done to optimize existing implementations of EOB for BBH systems [[EtienneMcWilliams]]. However, it is not clear that these optimizations can be readily adapted for the BNS model discussed here with dynamical tides. We therefore develop a surrogate model that accelerates the evaluation of EOB waveforms by interpolating between a sparsely sampled training set of waveforms within the parameter space.

%Surrogate modeling techniques have had significant success in gravitational-wave data analysis for accelerating the evaluation of a waveform $h(t;\bx)$ given its waveform parameters $\bx$. The standard technique is to construct a training set of waveforms that sample the parameter space, then compress this training set by building a reduced, orthonormal basis whose linear combination can reproduce any waveform in the parameter space to some desired accuracy. This serves to interpolate the waveform as a function of time (or frequency for frequency-domain waveforms) in terms of a reduced set of coefficients corresponding to the reduced set of basis functions. In previous work, this reduced basis has been constructed either with singular value decomposition (SVD) [[Michael's papers]] or with a greedy method that chooses a subset of the training-set waveforms that are as orthogonal to each other as possible, then orthonormalizes this subset with Graham-Schmidt orthogonalization [[All the other papers]].

Once this reduced basis is built, one can interpolate each of the coefficients $c_i(\bx)$ as a function of the waveform parameters $\bx$. This method has been used for frequency-domain surrogates of aligned spin BBH waveforms [[Michael's papers]]. Alternatively, one can use the empirical interpolation method (EIM) [[cite]] which re-expresses the linear combination of orthonormal basis functions in terms of a sum of empirical interpolating functions $B_j(t)$ times the value of the waveform $h(T_j; \bx)$ at empirical nodes $T_j$. One then interpolates the waveform $h(T_j; \bx)$ as a function of $\bx$ at each node $T_j$. This method has been used for time domain-waveforms for nonspinning [[cite]], aligned-spin [[cite]], and precessing [[cite]] BBH waveforms as well as nonspinning BNS waveforms with tidal interactions [[cite]], and frequency-domain, precessing waveforms [[cite]].

Previous work on surrogate models has focused mainly on parameter spaces $\bx$ with $\le 3$~dimensions. This allows one to approximate the coefficients $c_i(\bx)$ or waveform values $h(T_j; \bx)$ as functions of $\bx$ using simple fitting functions [[cite]] or standard interpolation techniques such as tensor spline~\cite{Puerrer2014, Puerrer2015} or Chebyshev interpolation~\cite{LackeyBernuzziGalley2017}. These interpolation techniques typically require waveforms evaluated on a rectangular grid, and this suffers from the curse of dimensionality; for $N$ points per dimension $d$,  $N^d$ waveforms are needed. For the 5-dimensional problem here, this is somewhat unfeasable. Recent work on optimally choosing waveforms for both EOB~\cite{DoctorFarrHolz2017} and NR~\cite{BlackmanFieldScheel2017a, BlackmanFieldScheel2017b} waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.

In this paper we focus on breaking this curse of dimensionality by using an interpolation method known as Gaussian process regression (GPR)~\cite{RasmussenWilliams2006} that does not require a regular grid. Importantly, GPR also provides a convenient estimate of its own uncertainty. This allows us to iteratively add new points to the training set in such a way as to minimize the interpolation error over the parameter space for a given number of waveforms in the training set. Similar approaches have recently been used to optimally sample waveforms for a 2-dimensional aligned spin BBH surrogate using GPR~\cite{DoctorFarrHolz2017} as well as for a surrogate of eccentric BBH mergers~\cite{HuertaMooreKumar2017}. GPR has also been used recently in GW data analysis to marginalize over waveform uncertainties in parameter estimation~\cite{MooreGair2014, MooreBerryChua2016}.

In this paper, we will develop methods to construct a surrogate model for the inspiral through merger of an aligned spin BNS waveform with tidal interactions using the EOB tidal description described in Hinderer et al.~[[cite1, cite2]]. This work is a continuation of the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS waveform using a slightly different description for the point-particle and tidal pieces by Bernuzzi et al.~\cite{Bernuzzi:2014owa}. Whereas in Ref.~\cite{LackeyBernuzziGalley2017} we constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model which significantly speeds up parameter estimation by not requiring an online Fourier transform for each waveform evaluation. It also enables additional techniques such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015} to accelerate likelihood evaluation.

\red{[[Obligitory summary:]]}

@@

\red{[[In the introduction you should emphasize the parts of this paper that are new to the GW field: Applying GPR to a high dimensional parameter space, showing that you can choose several new training set samples then calculate them in parallel without reconstructing the reduced basis. Building a surrogate for the residual to reduce the required interpolation accuracy. The first BNS waveform that realistically tapers to zero amplitude. Active learning method that readily extends to higher dimensions. By attaching to TaylorF2, the surrogate is valid for all frequencies.]]}

\red{[[We aim to construct a sufficiently accurate waveform surrogate using as few waveform evaluations as possible, to show how one could use NR simulations/go to higher dimensions.]]}

%% ________________________________________________________
\section{Aligned-spin, dynamical tides EOB model}

\red{[[Say somewhere that this model is based off of what is known as SEOBNRv4 (and/or 2?) in lalsuite.]]}

\red{[[The inspiral part is exactly as described in previous papers. The post-merger part is new to this paper. If this is true, we should write that somewhere.]]}

@@

We note that, to date, this is the only tidal EOB model that includes the effect of spin \red{[[Sebastiano also has a model that includes spin and tides.]]}\blue{[[Which is the Reference?]]}. More precisely, spin effects are those of the point-mass part of the model~\cite{Bohe:2016gbl}, that was calibrated to 141 NR simulations of aligned-spins BBHs.\footnote{This model is implemented in the LIGO Algorithm Library as SEOBNRv4.} The model, however, does not account for the quadrupole that the neutron stars acquire when they are spinning. This is a genuine matter effect that appears at lower PN order as compared to the tidal effects that are currently modeled. The importance of this physical effect in the context of data analysis for ground-based GW interferometers is currently being studied~\cite{IanTanja}.

@@

Dynamical tidal effects are implemented in the EOB model through a modification of the potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime. \red{[[Show how $\Delta_u$ enters the Hamiltonian.]]}\blue{[[There's no easy/illuminating way to explain how this potential enters the Hamiltonian]]}. In this paper we adopt the tidally-augmented expression of $\Delta_u$ that is discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}, namely
\begin{equation}
\Delta_u = \Delta_u^{\textrm{pm}} + \Delta_u^{\textrm{DT}}\,,
\end{equation}
where $\Delta_u^{\textrm{pm}}$ is the 4PN-accurate point-mass term of the underlying BBH EOB model of Ref.~\cite{Bohe:2016gbl} (see Eq.~(2.2) therein) and $\Delta_u^{\textrm{DT}}$ is the contribution due to dynamical tides. We consider the general case of a binary that may consist of two neutron stars; if either component of the binary is a black hole, then vanishing values of the tidal polarizabilities are employed in the formulas below. Let $A$ and $B$ be the body labels. Let $M_{A,B}$ be the mass of either body and $M=M_A+M_B$ be the total mass of the binary.  Explicitly, including quadrupolar and octupolar tidal effects, we have
\begin{widetext}
\begin{align}
 \Delta_u^{\textrm{DT}} =&  - 3\,\Lambda_{2,\textrm{dyn}}^{A}(u)X_A^4 X_B\,u^6  \left[ 1 + \frac{5}{2} X_Au + \left(3 + \frac{1}{8}X_A + \frac{337}{28} X_A^2\right)u^2 \right]\nonumber\\
 &- 15\,\Lambda_{3,\textrm{dyn}}^{A}(u)X_A^6X_B\,u^8 \left[1 + \left(-2 + \frac{15}{2}X_A\right)u + \left(\frac{8}{3} - \frac{311}{24}X_A + \frac{110}{3}X_A^2\right) u^2\right] + (A \leftrightarrow B)\,,
\end{align}
\end{widetext}
where $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate $r$ and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities. Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies $\hat{\omega}_{0\ell}^{A,B}$. In particular, the dimensionless dynamical tidal polarizability reads
\begin{equation}
\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\Lambda_{\ell}^{A,B}\hat{k}_{\ell\,\textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})\,,
\end{equation}
where $\Lambda_{\ell}^{A,B}$ is the dimensionless adiabatic tidal polarizability
\begin{equation}
\Lambda_{\ell}^{A,B}=\frac{2}{(2\ell-1)!!}\frac{k^{A,B}_{\ell}}{C_{A,B}^{2\ell+1}}\,,
\end{equation}
with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/M_{A,B}$ the neutron star compactness, which depends on the areal radius $R_{A,B}$ of the neutron star. Here $\hat{k}_{\ell\,\textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor given in Eq.~(11) of Ref.~\cite{Dietrich:2017feu}, which depends on the value of the $f$-mode $2^\ell$-pole dimensionless angular frequency, $\hat{\omega}_{0\ell}^{A,B}$.\footnote{Here $\hat{\omega}_{0\ell}^{A,B}=M_{A,B}\omega_{0\ell}^{A,B}$, where $\omega_{0\ell}^{A,B}$ is the frequency in geometrized units.} This enhancement factor is 1 for low orbital frequencies, and increases to about 2 as $\ell$ times the orbital frequency approaches the $f$-mode frequency.

@@

%%% ________________________________________________________
%\section{Aligned-spin, dynamical tides EOB model}
%Ever since its inception~\cite{Buonanno:1998gg}, the effective-one-body (EOB) approach to the general-relativistic 2-body problem has proven successful in modeling the dynamics and GW emission of binaries of compact objects. State-of-the-art aligned-spin EOB models~\cite{Bohe:2016gbl,Nagar:2017jdw} can accurately match hundreds of numerical-relativity (NR) simulations of spinning, nonprecessing binary black holes (BBHs) for mass ratios up to 8 and spin magnitudes up to 0.85 for unequal-mass (up to almost extremal for equal-mass) binaries. The EOB framework can also accommodate precessing-spin BBHs, showing good agreement to mildly precessing NR simulations~\cite{Babak:2016tgq}. This research program has been extended to accommodate tidal effects~\cite{Damour:2009wj,Vines:2010ca,Damour:2012yf,Bini:2012gu,Bernuzzi:2014owa,Hinderer:2016eia,Steinhoff:2016rfi,Dietrich:2017feu}, which become especially relevant in the strong-field regime for 2-body systems that contain matter --- typically binary neutron stars (BNSs) or neutron-star/black hole binaries (NSBHs). The main effect of tides is to make the gravitational interaction more attractive with respect to the vacuum case. In particular, Refs.~\cite{Hinderer:2016eia,Steinhoff:2016rfi} built upon the aligned-spin BBH EOB model of Ref.~\cite{Taracchini:2013rva} and proposed a way to include the effect of dynamical tides. Neutron stars that are part of a compact-object binary will deform in the tidal field generated by the companion --- be it a black hole or another neutron star. Because of binary motion, the forcing tidal field varies at the orbital frequency. Thus, in the late stages of the inspiral, the characteristic $f$-mode frequency of the neutron star can be dynamically approached, resulting in a resonant excitation of such mode. The net effect is an amplification of tidal effects as compared to the adiabatic limit, which assumes that the $f$-mode frequency is much larger than the frequency of the forcing tidal field. Dynamical tidal effects are implemented in the EOB model through a modification of the radial potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime. 
%
%In this paper we adopt the tidally-augmented expression of $\Delta_u$ that is discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}, namely
%\begin{equation}
%\Delta_u = \Delta_u^{\textrm{pm}} + \Delta_u^{\textrm{DT}}\,,
%\end{equation}

>>>>

@@

\pacs{
  % 04.25.D-,   % numerical relativity
  % 
  04.30.Db,   % gravitational wave generation and sources
  04.40.Dg,   % Relativistic stars: structure, stability, and oscillations
  % 04.70.Bw,   % classical black holes
  95.30.Sf,     % relativity and gravitation
  % 
  % 95.30.Lz,   % Hydrodynamics
  % 97.60.Jd    % Neutron stars
  % 97.60.Lf    % black holes (astrophysics)
  % 98.62.Mw    % Infall, accretion, and accretion disks
}

@@

The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} by the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors, has demonstrated that the parameters of the binary can be measured with enough precision to answer fundamental physics questions such as identifying BNS mergers as a source of gamma ray bursts [[cite]], measuring the Hubble constant~\cite{GW170817Hubble}, and understanding the structure of neutron stars (NSs) through tidal interactions~\cite{GW170817}. Measuring these parameters accurately will also be necessary to properly interpret the gamma ray burst [[cite]] and optical afterglow [[cite]] that was found in conjunction. Furthermore, an estimate of the BNS merger rate of 320--4740 events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval)~\cite{GW170817} indicates that, when second generation detectors reach design sensitivity, including KAGRA~\cite{Somiya2012} and possibly LIGO-India~\cite{IyerSouradeepUnnikrishnan2011}, we could eventually observe a population of tens or hundreds of events after a couple of years of observation~\cite{LIGORate2010}. This will allow us to estimate the NS mass distribution [[cite]], spin distribution [[cite]], and universal equation of state (EOS) that is common to all NSs\cite{DelPozzoLiAgathos2013, LackeyWade2015}.

These results are obtained using Bayesian analyses that compare the data to parameterized waveform models. To avoid systematic errors highly accurate models are needed, and they must include all relevant physical effects. This includes descriptions of the point-particle interactions, spins, and tides. Prior to the discovery of GW170817, most parameter estimation analyses were done using post-Newtonian (PN) waveform models. These models are currently known up to 3.5 PN order in the point-particle and spin terms, and the leading and next-to-leading order terms for tidal interactions that enter at 5 and 6PN order. Unfortunately, the lack of point-particle terms above 3.5PN order leads to systematic errors in the tidal parameters that are as large as the statistical errors [[Favata, YagiYunes, Wade]]. Alternative models exist with more accurate representations of the point particle terms. However, they do not always simultaneously include the spin and tidal effects as well. Not including spin effects, for example, can bias the measured tidal parameter for even moderate dimensionless spins of $|\chi_i| \le 0.04$ [[Favata]]. In addition, ignoring tidal interactions can bias the measured mass ratio from equal mass to unequal mass [[cite who?]].

One such alternative to the PN waveforms is the effective one body (EOB) formalism (discussed in section~\ref{sec:eob}). It includes several dynamical effects beyond the standard PN adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations. Until recently, however, no version of the EOB formalism included the effects of spin and tides in the same model. Recently two versions have become available that include both tidal effects and aligned spin. These models are being implemented within the LIGO Algorithm Library (LAL) [[cite]] module \texttt{lalsimulation} under the names \texttt{SEOBNRv4T} and \texttt{TEOBResumS}. The \texttt{SEOBNRv4T} version [[cite]] used in this paper, includes the tidally induced $\ell=2$ and 3  multipole moments as well as the induced $\ell=2$ and 3 $f$-mode resonances, but does not yet include the spin induced quadrupole moment that can be important for large spins~\cite{Poisson, HarryHinderer2017}. This model also agrees with NR simulations of BNS systems to a level consistent with the numerical accuracy of the simulations, which is less than 1~rad during the last $\sim 10$ orbits before merger~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}. The alternative version, \texttt{TEOBResumS} [[cite]], uses a slightly different point-particle and spin prescription. And, in addition, includes the $\ell = 2, 3$, and 4 tidal effects and the spin induced quadrupole moment, but does not include the effect of dynamical tides from the induced $f$-mode resonances. It does, however, use a method for re-summing the expressions for tidal interactions. It also shows similar consistency with NR simulations [[Is this true? cite]].

In addition to accuracy, useful waveform models must be fast enough to be used in Bayesian parameter estimation algorithms such as Markov chain Monte Carlo (MCMC) and Nested Sampling that currently require $10^7$--$10^8$ sequential waveform evaluations. Waveform evaluation times must therefore be significantly less than 1~s in order to run in less than a month for each event. Most time-domain waveform implementations such as the PN approximants and the EOB models involve solving a set of ordinary differential equations and then Fourier transforming the results into the frequency domain. For PN waveform models, this can take several seconds to minutes, while for the \texttt{SEOBNRv4T} model discussed here, evaluation time can be more than a day for a starting frequency of 10~Hz. Significant work has been done to optimize existing implementations of EOB for BBH systems~\cite{DevineEtienneMcWilliams2016}. However, it is not clear that sufficient speed ups can be achieved for the BNS model discussed here with dynamical tides.

For the GW170817 event, therefore, a set of frequency-domain analytic or semi-analytic waveform models were used as templates. The first was the stationary phase approximation PN waveform model known as TaylorF2 [[cite]]. Because of the known systematic errors that result from PN waveforms, two additional models were produced for comparison: \texttt{IMRPhenomD\_NRTides} and \texttt{SEOBNRv4\_ROM\_NRTides}. Both of these models used an identical prescription for tides presented in Ref.~\cite{DietrichBernuzziTichy2017}; at low frequencies, the tidal effect is that of TaylorF2. It then transitions to a fit to the tidal effect measured by a small set of numerical simulations with varying mass ratio and aligned-spins. The point-particle and spin prescriptions were different. For \texttt{IMRPhenomD\_NRTides}, the waveform is based on the phenomenological \texttt{IMRPhenomD} BBH model [[cite]]. It uses TaylorF2 at low frequencies then transitions to fits to the EOB waveform at mid frequencies and fits to NR simulations at high frequencies. The \texttt{SEOBNRv4\_ROM\_NRTides} model is based on the fast surrogate model \texttt{SEOBNRv4\_ROM} [[cite]] to the EOB waveform \texttt{SEOBNRv4} for BBH systems [[cite]]. Both of these models are improvements to \texttt{TaylorF2}, but do not incorporate all known physical effects in a consistent way. We therefore build a fast surrogate that reproduces as accurately as possible \texttt{SEOBNRv4T} in the frequency domain.

Surrogate modeling techniques have had significant success in gravitational-wave data analysis for accelerating the evaluation of waveform models. The essential idea is to construct an interpolating function that approximates a waveform model as a function of frequency (or time) and waveform parameters $\bx$. Previous works have focused on efficiently representing the parameter space of waveforms using a reduced basis of orthonormal functions whose linear combination can reproduce any waveform in the parameter space to some desired accuracy. The two most common approaches have been singular value decomposition (SVD)~\cite{Puerrer2014, Puerrer2015} or a greedy method that selects a subset of the training-set waveforms then orthonormalizes them with the Graham-Schmidt process~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}. In this work, instead of using a reduced basis to efficiently represent the EOB waveform as a function of frequency, we will instead use the fact that the frequency-domain waveform is approximately known analytically via the TaylorF2 waveform. We then build a surrogate of the difference (or residual) between the EOB waveform and the TaylorF2 waveform. Because this residual is small, extremely high accuracy is not required, and we simply choose to represent the residual as a function of frequency using cubic splines between a small set of frequency nodes.

This solves the problem of efficiently representing waveforms as a function of frequency, but leaves the question of interpolating as a function of waveform parameters $\bx$ unanswered. Previous work on surrogate models have focused mainly on parameter spaces with three or fewer dimensions. This allows one to approximate the waveform as a function of $\bx$ using standard interpolation techniques such as tensor spline~\cite{Puerrer2014, Puerrer2015} or Chebyshev interpolation~\cite{LackeyBernuzziGalley2017}. These interpolation techniques typically require waveforms to be evaluated on a rectangular grid, and this suffers from the curse of dimensionality; for $N$ points per dimension $d$,  $N^d$ waveforms are needed. For the problem here, we will only be able to reduce the parameter space to five dimensions, so grid based interpolation will be unfeasable. Recent work on optimally choosing waveforms for both EOB~\cite{DoctorFarrHolz2017} and NR~\cite{BlackmanFieldScheel2017a, BlackmanFieldScheel2017b} waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.

In this paper we focus on an interpolation method known as Gaussian process regression (GPR)~\cite{RasmussenWilliams2006} that does not require a regular grid. Importantly, GPR also provides a convenient estimate of its own uncertainty. This allows us to iteratively add new points to the training set in such a way as to minimize the interpolation error over the parameter space for a given number of waveforms in the training set. Similar approaches have recently been used to optimally sample waveforms for a 2-dimensional aligned spin BBH surrogate using GPR~\cite{DoctorFarrHolz2017} as well as for a surrogate of eccentric BBH mergers~\cite{HuertaMooreKumar2017}. GPR has also been used in GW data analysis to marginalize over waveform uncertainties in parameter estimation~\cite{MooreGair2014, MooreBerryChua2016} and to estimate noise power spectral densities [[cite]].

This work is a continuation of the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform using the description for the point-particle and tidal terms by Bernuzzi et al.~\cite{Bernuzzi:2014owa}. Whereas in Ref.~\cite{LackeyBernuzziGalley2017} we constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model which significantly speeds up parameter estimation by not requiring an online Fourier transform for each waveform evaluation. It also enables additional techniques such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015} and multi-banding~\cite{VinciguerraVeitchMandel2017} to accelerate likelihood evaluation. The resulting surrogate is the first EOB waveform model that consistently incorporates spin and tides and is fast enough for production parameter estimation codes.

@@

%% ________________________________________________________
\section{Aligned-spin, dynamical tides EOB model}
\label{sec:eob}

@@

%We note that, to date, this is the only tidal EOB model that includes the effect of spin \red{[[Sebastiano also has a model that includes spin and tides.]]}\blue{[[Which is the Reference?]]}.

More precisely, spin effects are those of the point-mass part of the model~\cite{Bohe:2016gbl}, that was calibrated to 141 NR simulations of aligned-spins BBHs.\footnote{This model is implemented in the LIGO Algorithm Library as SEOBNRv4.} The model, however, does not account for the quadrupole that the neutron stars acquire when they are spinning. This is a genuine matter effect that appears at lower PN order as compared to the tidal effects that are currently modeled. The importance of this physical effect in the context of data analysis for ground-based GW interferometers is currently being studied~\cite{IanTanja}.

@@

Dynamical tidal effects are implemented in the EOB model through a modification of the potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime. In this paper we adopt the tidally-augmented expression of $\Delta_u$ that is discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}, namely
\begin{equation}
\Delta_u = \Delta_u^{\textrm{pm}} + \Delta_u^{\textrm{DT}}\,,
\end{equation}
where $\Delta_u^{\textrm{pm}}$ is the 4PN-accurate point-mass term of the underlying BBH EOB model of Ref.~\cite{Bohe:2016gbl} (see Eq.~(2.2) therein) and $\Delta_u^{\textrm{DT}}$ is the contribution due to dynamical tides. We consider the general case of a binary that may consist of two neutron stars; if either component of the binary is a black hole, then vanishing values of the tidal polarizabilities are employed in the formulas below. Let $A$ and $B$ be the body labels. Let $M_{A,B}$ be the mass of either body and $M=M_A+M_B$ be the total mass of the binary.  Explicitly, including quadrupolar and octupolar tidal effects, we have
\begin{widetext}
\begin{align}
 \Delta_u^{\textrm{DT}} =&  - 3\,\Lambda_{2,\textrm{dyn}}^{A}(u)X_A^4 X_B\,u^6  \left[ 1 + \frac{5}{2} X_Au + \left(3 + \frac{1}{8}X_A + \frac{337}{28} X_A^2\right)u^2 \right]\nonumber\\
 &- 15\,\Lambda_{3,\textrm{dyn}}^{A}(u)X_A^6X_B\,u^8 \left[1 + \left(-2 + \frac{15}{2}X_A\right)u + \left(\frac{8}{3} - \frac{311}{24}X_A + \frac{110}{3}X_A^2\right) u^2\right] + (A \leftrightarrow B)\,,
\end{align}
\end{widetext}
where $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate $r$ and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities. Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies $\hat{\omega}_{0\ell}^{A,B}$. In particular, the dimensionless dynamical tidal polarizability reads
\begin{equation}
\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\Lambda_{\ell}^{A,B}\hat{k}_{\ell\,\textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})\,,
\end{equation}
where $\Lambda_{\ell}^{A,B}$ is the dimensionless adiabatic tidal polarizability
\begin{equation}
\Lambda_{\ell}^{A,B}=\frac{2}{(2\ell-1)!!}\frac{k^{A,B}_{\ell}}{C_{A,B}^{2\ell+1}}\,,
\end{equation}
with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/M_{A,B}$ the neutron star compactness, which depends on the areal radius $R_{A,B}$ of the neutron star. Here $\hat{k}_{\ell\,\textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor given in Eq.~(11) of Ref.~\cite{Dietrich:2017feu}, which depends on the value of the $f$-mode $2^\ell$-pole dimensionless angular frequency, $\hat{\omega}_{0\ell}^{A,B}$.\footnote{Here $\hat{\omega}_{0\ell}^{A,B}=M_{A,B}\omega_{0\ell}^{A,B}$, where $\omega_{0\ell}^{A,B}$ is the frequency in geometrized units.} This enhancement factor is 1 for low orbital frequencies, and increases to about 2 as $\ell$ times the orbital frequency approaches the $f$-mode frequency.

@@
